---
title: "Analysis of Dorsal Targets in Raf LOF Mutation"
output: 
  html_document:
    toc: true
    toc_float: true
    df_print: paged
---

```{r setup, include=FALSE}
short=TRUE #if short==TRUE, do not echo code chunks
debug=FALSE
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
knitr::opts_chunk$set(fig.width=12, fig.height=8, fig.path='figures/vis_enrichments/',
                      echo=!short, warning=debug, message=debug, dev=c("png", "pdf"))
```

```{r loading libraries, include=FALSE}
source("analysis/functions.R")
library(DESeq2)
library(knitr)
library(GenomicFeatures)
library(RColorBrewer)
library(grid)
library(gridExtra)
library(lattice)
library(reshape)
library(geneplotter)
library(pheatmap)
library(Biostrings)
library(BSgenome.Dmelanogaster.UCSC.dm6)
library(clusterProfiler)
library(stringr)
library(tidyverse)
library(DT)
library(data.table)
library(ggpubr)
library(cowplot)
library(ggrepel)
library(msigdbr)
options(connectionObserver = NULL)
#BiocManager::install("org.Dm.eg.db", character.only = TRUE)
library("org.Dm.eg.db", character.only = TRUE)
library(pathview)
library(enrichR)
library(VennDiagram)
library(png)
```

```{r}
## If the directory doesn't exist, create it
if(!dir.exists("analysis/output/V_E")){
  dir.create("analysis/output/V_E")
}
```

```{r loading in data, include=FALSE}
## Outputs required from QC.Rmd
experimental_metadata <- readRDS("analysis/output/QC/experimental_metadata.rds")
tpm <- readRDS("analysis/output/QC/tpm.rds")
ensembl_genes <- readRDS("analysis/output/QC/ensembl_genes.rds")

## Outputs required from DE.Rmd
dds <- readRDS("analysis/output/DE/dds_de.rds")
significant_rld <- readRDS("analysis/output/DE/significant_rld.rds")
wald_sig <- readRDS("analysis/output/DE/wald_sig_results.rds")
wald_all <- readRDS("analysis/output/DE/wald_all_results.rds")

## Global Variables
options(ggrepel.max.overlaps = Inf) ##ggrepel option
nc <- counts(dds, normalized = TRUE) ##normalised counts
lfc.threshold = log2(1.5) ##log2fc threshold
wald_all$entrez_id = ensembl_genes$entrezgene_id[match(row.names(wald_all), ensembl_genes$gene_id)]
wald_all_df <- wald_all %>%
  as.data.frame()
wald_sig$entrez_id = ensembl_genes$entrezgene_id[match(row.names(wald_sig), ensembl_genes$gene_id)]
wald_sig_df <- as.data.frame(wald_sig)
```

```{r z-score transformation}
# assay(x) to access the count data
sig_mat_rld = assay(significant_rld)

# The apply function swaps the rows to samples and columns to genes -- the standard is the other way around: samples in cols and genes in rows, hence the transpose function
zscores = t(apply(sig_mat_rld, 1, function(x){ (x - mean(x)) / sd(x) }))
```

```{r heatmap annotation}
## >> !! NOTE !! The annotation needs to be in the correct order as the condition in the metadata.txt
annotation = data.frame(Condition = rep(c("Control", "Mutant_LOF"),
                                            c(3, 3)))

row.names(annotation) = experimental_metadata$sample_id 
anno_colours = list(Condition = c(Control = "#72147e", Mutant_LOF= "#fea82f"), 
                    Germ_Layer = c(Ectoderm = "#45526c", Mesoderm = "#5aa897", Neuroectoderm = "#d2e69c"))
```

## Heatmap
```{r}
oe <-
  wald_sig %>% 
  as.data.frame() %>%
  filter(log2FoldChange > 0) %>%
  filter(!(is.na(external_gene_name))) %>%
  arrange(padj)

ue <-
  wald_sig %>% 
  as.data.frame() %>%
  filter(log2FoldChange < 0) %>%
  filter(!(is.na(external_gene_name))) %>%
  arrange(padj)

sig_genes <- rbind(oe, ue)

## So that the zscores are arranged in the order of oe and ue
zscores= zscores[rownames(sig_genes),]

pheatmap(zscores,
         color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(100),
         annotation = annotation,
         annotation_colors = anno_colours,
         cluster_rows = F, 
         cluster_cols = F)
```

## Number of Genes

```{r extracting genes}
data.frame(Cluster = c("Upregulated_MUT", "Downregulated_MUT", "Total"),
                        Number_of_genes = c(nrow(oe),
                                            nrow(ue),
                                            (nrow(oe)+nrow(ue))))
```

## Raf
```{r}
nc_df = as.data.frame(nc)
nc_df$gene_names = ensembl_genes$external_gene_name[match(rownames(nc_df), ensembl_genes$gene_id)]


nc_df["FBgn0003079",] %>%
    melt() %>%
    select(variable,value) %>%
    `colnames<-`(c("Sample", "Value")) %>%
    mutate(Condition = ifelse(grepl("WT", Sample), "WT", "LOF_Mutant")) %>%
    ggplot(aes(x=Condition, y=Value, fill = Condition)) + 
    geom_boxplot() + 
    theme_classic() + 
    theme(axis.text.x = element_blank(), legend.position="none") +
    labs(title = "Raf") + 
    xlab(NULL) + 
    ylab(NULL) +
    scale_x_discrete(limits = rev) +
    expand_limits(y=0) +
    scale_y_continuous(limits = c(1000, 3000), breaks = seq(1000,3000,by = (3000-1000)/5))

```

## Volcano Plot of Significant Genes
```{r}
make_volcano_with_labels(wald_sig, "", "Volcano Plot: S926glc(Mutant_LOF) vs. WT")
```

## Volcano Plot of All Expressed Genes
```{r}
make_volcano_with_labels(wald_all, "", "Volcano Plot: S926glc(Mutant_LOF) vs. WT")
```

## Enrichments with EnrichR

```{r}
setEnrichrSite("FlyEnrichr")
dbs <- listEnrichrDbs()
to_check <- c("GO_Biological_Process_2018", "KEGG_2019")
```

### Upregulated in Mutant
```{r}
oe_enrichr <- oe$external_gene_name
eresList <- enrichr(oe_enrichr, databases = to_check)

eresList$GO_Biological_Process_2018  %>% 
  plot_enrichr(name = "GO_Biological_Process_2018")
eresList$GO_Biological_Process_2018  %>% 
  datatable(options = list(scrollX = TRUE), class = "white-space: nowrap")
eresList$GO_Biological_Process_2018 %>%
  write.csv("analysis/output/V_E/upregulated_go.csv", row.names = TRUE, col.names = TRUE)

eresList$KEGG_2019  %>% 
  plot_enrichr(name = "KEGG_2019")
eresList$KEGG_2019  %>% 
  datatable(options = list(scrollX = TRUE), class = "white-space: nowrap")
eresList$KEGG_2019 %>%
  write.csv("analysis/output/V_E/upregulated_kegg.csv", row.names = TRUE, col.names = TRUE)
```

### Downregulated in Mutant
```{r}
ue_enrichr <- ue$external_gene_name
eresList <- enrichr(ue_enrichr, databases = to_check)

eresList$GO_Biological_Process_2018  %>% 
  plot_enrichr(name = "GO_Biological_Process_2018")
eresList$GO_Biological_Process_2018  %>% 
  datatable(options = list(scrollX = TRUE), class = "white-space: nowrap")
eresList$GO_Biological_Process_2018 %>%
  write.csv("analysis/output/V_E/downregulated_go.csv", row.names = TRUE, col.names = TRUE)


eresList$KEGG_2019  %>% 
  plot_enrichr(name = "KEGG_2019")
eresList$KEGG_2019  %>% 
  datatable(options = list(scrollX = TRUE), class = "white-space: nowrap")
eresList$KEGG_2019 %>%
  write.csv("analysis/output/V_E/downregulated_kegg.csv", row.names = TRUE, col.names = TRUE)
```


## GSEA Analysis
### clusterProfiler
#### Using MSigDB C5
```{r}
## Use the MSigDB R package to download C5 -- Ontology -- gene sets for Drosophila. Check MSigDB websites for other categories.
gene_sets <- msigdbr(species = "Drosophila melanogaster", category = "C5")
gene_sets <- gene_sets %>%
  dplyr::select(gs_name, gene_symbol)

## 1. Rank genes in results
gsea_prep <- wald_all_df %>%
  arrange(padj) %>%
  mutate(padj = case_when(padj == 0 ~ .Machine$double.xmin, TRUE ~ padj)) %>%
  mutate(gsea_metric = -log10(padj) * sign(log2FoldChange)) %>%
  filter(!is.na(gsea_metric)) %>%
  arrange(desc(gsea_metric))

## 2. Get vector of gene names with their ranking
gsea_ranks <- gsea_prep %>% 
  dplyr::select(external_gene_name, gsea_metric) %>%
  distinct(external_gene_name, .keep_all = TRUE) %>%
  deframe()

## 3. Run GSEA to get enrichments
gsea_res <- GSEA(geneList = gsea_ranks,
                TERM2GENE = gene_sets)

gsea_res %>% 
  data.frame() %>%
  datatable(options = list(scrollX = TRUE), class = "white-space: nowrap")

#gseaplot(gsea_res,"GO_DORSAL_VENTRAL_PATTERN_FORMATION", title = "Dorsal Ventral Pattern Formation")
```

## Pathway Visualisation
```{r}
## getting kegg enrichments
kegg = enrichKEGG(gene = wald_sig$entrez_id, 
                     organism = "dme",
                     keyType="ncbi-geneid", 
                     universe=as.character(na.omit(ensembl_genes[rownames(counts(dds))]$entrezgene_id)))

kegg_df <- kegg@result %>%
  data.frame()
```

```{r}
wald_fc = wald_all$log2FoldChange
names(wald_fc) = wald_all$entrez_id

## extracting the pathway IDs of interest
pathways_interest = data.frame()

toll = kegg_df %>%
  filter(grepl("Toll",Description, fixed = TRUE))

mapk = kegg_df %>%
  filter(grepl("MAPK",Description))

pathways_interest = rbindlist(list(pathways_interest, toll, mapk))

pid = pathways_interest$ID

## This function annoyingly prints out png files to the main directory
plot_pathview = function(pid){pathview(gene.data=wald_fc, pathway.id=pid, species="dme", new.signature=FALSE, limit=list(gene=2, cpd=1), low = list(gene = "#4575b4", cpd = "#4575b4"), mid = list(gene = "white", cpd = "gray"), high = list(gene = "#d73027", cpd = "yellow"))}

## Applying the function above to pathway IDs. Assigning it to random variable because it returns an empty object. 
tmp = sapply(pid, plot_pathview)
```

### Toll Signalling Pathway
```{r}
toll_png <- readPNG("dme04624.pathview.png")
p <- rasterGrob(toll_png, interpolate = TRUE)
ggarrange(p)
```

### MAPK Signalling Pathway
```{r}
mapk_png <- readPNG("dme04013.pathview.png")
p <- rasterGrob(mapk_png, interpolate = TRUE)
ggarrange(p)
```

## Dorsal Target Genes
### Dorsal Targets in the Mesoderm
```{r}
known_meso_targets = c("FBgn0000719", "FBgn0010389", "FBgn0011656", "FBgn0003448", "FBgn0004110", "FBgn0003900", "FBgn0004606")
mi_meso_targets = c("FBgn0014380", "FBgn0037207", "FBgn0044049", "FBgn0034726", "FBgn0004512", "FBgn0000044", "FBgn0000116", "FBgn0034075", "FBgn0032693", "FBgn0014179", "FBgn0286786", "FBgn0003308", "FBgn0020299","FBgn0262139")

all_meso_targets = c(known_meso_targets, mi_meso_targets)
meso_gene_names = sig_genes[rownames(sig_genes) %in% all_meso_targets,] 

zscores_meso = get_zscores(zscores, meso_gene_names)

(pheatmap(zscores_meso,
           color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(100),
           fontsize_row = 10,
           annotation_col = annotation,
           annotation_colors = anno_colours,
           cluster_rows = FALSE,
           cluster_cols = FALSE,
           border_color = NA))

```

```{r fig.width=15,fig.height=15}
nc_meso = get_df(all_meso_targets)
nc_meso = nc_meso[order(row.names(nc_meso)),]

legend <- extracting_legend(nc_meso, 1)

## use rownames(nc_meso) to check which names to use
act57b <- gene_boxplot(nc_meso, "Act57B", 0, 100000)
argk <- gene_boxplot(nc_meso, "Argk", 0, 15000)
asph <- gene_boxplot(nc_meso, "Asph", 0, 5000)
cyp <- gene_boxplot(nc_meso, "Cyp310a1", 0, 1200)
fog <- gene_boxplot(nc_meso, "fog", 2000, 4000)
gcm <- gene_boxplot(nc_meso, "gcm", 0, 2000)
htl <- gene_boxplot(nc_meso, "htl", 0, 1750)
ilp <- gene_boxplot(nc_meso, "Ilp4", 0, 4000)
mdr <- gene_boxplot(nc_meso, "Mdr49", 500, 2000)
mef <- gene_boxplot(nc_meso, "Mef2", 0000, 8000)
mes2 <- gene_boxplot(nc_meso, "Mes2", 2000, 4000)
mes4 <- gene_boxplot(nc_meso, "Mes4", 200, 600)
rhol <- gene_boxplot(nc_meso, "RhoL", 400, 1000)
ry <- gene_boxplot(nc_meso, "ry", 500, 1750)
sna <- gene_boxplot(nc_meso, "sna", 0, 3000)
stumps <- gene_boxplot(nc_meso, "stumps", 0, 6000)
tin <- gene_boxplot(nc_meso, "tin", 0, 1000)
trh <- gene_boxplot(nc_meso, "trh", 0, 3500)
twi <- gene_boxplot(nc_meso, "twi", 0, 3000)
zfh <- gene_boxplot(nc_meso, "zfh1", 500, 5000)
 
all_plots <- plot_grid(plotlist = list(act57b, argk, asph, cyp,fog, gcm,htl,ilp,mdr,mef,mes2,mes4,rhol,ry,sna,stumps,tin,trh,twi,zfh), nrow = ceiling(length(rownames(nc_meso))/5), ncol = 5)
plot_grid(legend, all_plots, nrow = 2, ncol = 1, rel_heights = c(.1, 1))
```

```{r}
make_volcano_with_labels(wald_all, all_meso_targets, title = "Dorsal Targets in the Mesoderm")
```

#### Hypergeometric Test
```{r, echo=TRUE}
q = nrow(zscores_meso)
m = nrow(zscores_meso)
n = nrow(counts(dds)) - nrow(zscores_meso)
k = nrow(ue)
  
phyper(q, m, n, k, lower.tail = FALSE)
```


### Dorsal Targets in the Neuroectoderm
```{r}
known_neuroecto_targets = c("FBgn0024250", "FBgn0025776", "FBgn0002561", "FBgn0004635", "FBgn0004666", "FBgn0003463", "FBgn0261930")
mi_neuroecto_targets = c("FBgn0033159", "FBgn0037085", "FBgn0265140", "FBgn0033652", "FBgn0040271", "FBgn0000547", "FBgn0002543", "FBgn0003892", "FBgn0043575", "FBgn0003326", "FBgn0041184", "FBgn0029123", "FBgn0011739")

all_neuroecto_targets = c(known_neuroecto_targets, mi_neuroecto_targets)
neuroecto_gene_names = sig_genes[rownames(sig_genes) %in% all_neuroecto_targets,] 

zscores_neuroecto = get_zscores(zscores, neuroecto_gene_names)

(pheatmap(zscores_neuroecto,
           color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(100),
           fontsize_row = 10,
           annotation_col = annotation,
           annotation_colors = anno_colours,
           cluster_rows = FALSE,
           cluster_cols = FALSE,
           border_color = NA))
```

```{r fig.width=15,fig.height=15}
nc_neuroecto = get_df(all_neuroecto_targets)
nc_neuroecto = nc_neuroecto[order(row.names(nc_neuroecto)),]

legend <- extracting_legend(nc_neuroecto, 1)

brk <- gene_boxplot(nc_neuroecto, "brk", 0, 3000)
dscam <- gene_boxplot(nc_neuroecto, "Dscam1", 2000, 8000)
ed <- gene_boxplot(nc_neuroecto, "ed", 20000, 30000)
ind <- gene_boxplot(nc_neuroecto, "ind", 0, 350)
lsc <- gene_boxplot(nc_neuroecto, "l(1)sc", 0, 2500)
meltrin <- gene_boxplot(nc_neuroecto, "Meltrin", 2000, 4000)
neu <- gene_boxplot(nc_neuroecto, "Neu2", 0, 400)
pgrp <- gene_boxplot(nc_neuroecto, "PGRP-SC2", 0, 300)
ptc <- gene_boxplot(nc_neuroecto, "ptc", 1000, 10000)
rho <- gene_boxplot(nc_neuroecto, "rho", 0, 1000)
robo2 <- gene_boxplot(nc_neuroecto, "robo2", 5000, 20000)
sca <- gene_boxplot(nc_neuroecto, "sca", 0, 8000)
sim <- gene_boxplot(nc_neuroecto, "sim", 0, 500)
socs <- gene_boxplot(nc_neuroecto, "Socs36E", 1000, 5000)
sog <- gene_boxplot(nc_neuroecto, "sog", 0, 2500)
soxn <- gene_boxplot(nc_neuroecto, "SoxN", 0, 8000)
sulf <- gene_boxplot(nc_neuroecto, "Sulf1", 3000, 5000)
ths <- gene_boxplot(nc_neuroecto, "ths", 1000, 2500)
vnd <- gene_boxplot(nc_neuroecto, "vnd", 0, 800)
wts <- gene_boxplot(nc_neuroecto, "wts", 2000, 4000)

all_plots <- plot_grid(plotlist = list(brk,dscam,ed,ind,lsc,meltrin,neu,pgrp,ptc,rho,robo2,sca,sim,socs,sog,soxn,sulf,ths,vnd,wts), nrow = ceiling(length(rownames(nc_neuroecto))/5), ncol = 5)
plot_grid(legend, all_plots, nrow = 2, ncol = 1, rel_heights = c(.1, 1))
```

```{r}
make_volcano_with_labels(wald_all, all_neuroecto_targets, title = "Dorsal Targets in the Neuroectoderm")
```

#### Hypergeometric Test
```{r, echo=TRUE}
q = nrow(zscores_neuroecto)
m = nrow(zscores_neuroecto)
n = nrow(counts(dds)) - nrow(ue)
k = nrow(ue)
  
phyper(q, m, n, k, lower.tail = FALSE)
```

### Dorsal Targets in the Ectoderm
```{r}
# These are supposed to go up in mutant

known_ecto_targets = c("FBgn0000490", "FBgn0003117", "FBgn0012037", "FBgn0003719", "FBgn0003896", "FBgn0003963", "FBgn0004053")
mi_ecto_targets = c("FBgn0033483", "FBgn0028789", "FBgn0260746", "FBgn0262579", "FBgn0004863", "FBgn0014454", "FBgn0020637")

all_ecto_targets = c(known_ecto_targets, mi_ecto_targets)
ecto_gene_names = sig_genes[rownames(sig_genes) %in% all_ecto_targets,] 

zscores_ecto = get_zscores(zscores, ecto_gene_names)

(pheatmap(zscores_ecto,
           color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(100),
           fontsize_row = 10,
           annotation_col = annotation,
           annotation_colors = anno_colours,
           cluster_rows = FALSE,
           cluster_cols = FALSE,
           border_color = NA))
```

```{r}
nc_ecto = get_df(all_ecto_targets)
nc_ecto = nc_ecto[order(row.names(nc_ecto)),]

legend <- extracting_legend(nc_ecto, 1)

ance <- gene_boxplot(nc_ecto, "Ance", 5000, 15000)
c15 <- gene_boxplot(nc_ecto, "C15", 0, 2000)
doc1 <- gene_boxplot(nc_ecto, "Doc1", 0, 1500)
dpp <- gene_boxplot(nc_ecto, "dpp", 1000, 1600)
ect <- gene_boxplot(nc_ecto, "Ect3", 1500, 3000)
egr <- gene_boxplot(nc_ecto, "egr", 1000, 1600)
lcp <- gene_boxplot(nc_ecto, "Lcp65Ag2", 0, 60000)
pnr <- gene_boxplot(nc_ecto, "pnr", 1000, 5000)
sarm <- gene_boxplot(nc_ecto, "Sarm", 1500, 6500)
tld <- gene_boxplot(nc_ecto, "tld", 500, 1200)
tup <- gene_boxplot(nc_ecto, "tup", 2000, 6000)
ush <- gene_boxplot(nc_ecto, "ush", 2000, 6000)
zen <- gene_boxplot(nc_ecto, "zen", 0, 600)

all_plots <- plot_grid(plotlist = list(ance,c15,doc1,dpp,ect,egr,lcp,pnr,sarm,tld,tup,ush,zen), nrow = ceiling(length(rownames(nc_ecto))/5), ncol = 5)
plot_grid(legend, all_plots, nrow = 2, ncol = 1, rel_heights = c(.1, 1))
```

```{r}
make_volcano_with_labels(wald_all, all_ecto_targets, title = "Dorsal Targets in the Ectoderm")
```

#### Hypergeometric Test
```{r, echo=TRUE}
q = nrow(zscores_ecto) - 1
m = nrow(zscores_ecto)
n = nrow(counts(dds)) - nrow(oe)
k = nrow(oe)
  
phyper(q, m, n, k, lower.tail = FALSE)
```

### Combined Heatmap
```{r}
comb_meso = all_meso_targets[all_meso_targets %in% rownames(sig_genes)]
comb_neuroecto = all_neuroecto_targets[all_neuroecto_targets %in% rownames(sig_genes)]
comb_ecto = all_ecto_targets[all_ecto_targets %in% rownames(sig_genes)]
all_dorsal_targets = c(comb_meso, comb_neuroecto, comb_ecto)
all_gene_names = sig_genes[rownames(sig_genes) %in% all_dorsal_targets,]
all_gene_names$Germ_Layer = "Mesoderm"
all_gene_names[rownames(all_gene_names) %in% comb_neuroecto,]$Germ_Layer = "Neuroectoderm"
all_gene_names[rownames(all_gene_names) %in% comb_ecto,]$Germ_Layer = "Ectoderm"

zscores_all = get_zscores(zscores, all_gene_names)

anno_row <- all_gene_names %>%
  dplyr::select(external_gene_name, Germ_Layer) %>%
  remove_rownames() %>%
  column_to_rownames("external_gene_name") %>%
  arrange(Germ_Layer)

zscores_all = zscores_all[rownames(anno_row),]

(pheatmap(zscores_all,
           color = colorRampPalette(rev(brewer.pal(n = 7, name = "RdBu")))(100),
           fontsize_row = 10,
           annotation_col = annotation,
           annotation_colors = anno_colours,
           annotation_row = anno_row,
           cluster_rows = FALSE,
           cluster_cols = FALSE,
           border_color = NA))
```


```{r write out files}
## to be used in Koenecke Analysis
saveRDS(oe, "analysis/output/V_E/oe.rds")
saveRDS(ue, "analysis/output/V_E/ue.rds")
```


## Session Info
```{r}
sessionInfo()
```
